Ensemble Stability of Biological Networks

### 2014-10-07 paper.tex
reorder authors and update cited abstract
update notation with angle bracket averages

% Title must be 150 characters or less
\begin{center}
{\Large
\textbf{Hierarchical Network Structure Promotes Dynamical Robustness}
}
% Insert Author names, affiliations and corresponding author email.
\\[.5cm]
Aviv Bergman$^{1,2,3,4}$,
Raymond S. Puzio$^{1}$,
Cameron Smith$^{1}$,
\\[.5cm]
$^1$Department of Systems and Computational Biology,\\
$^2$Dominick P. Purpura Department of Neuroscience,\\
$^3$Department of Pathology, Albert Einstein College of Medicine,\\
1301 Morris Park Ave, Bronx, NY 10461, USA\\
$^4$Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA
\\[.5cm]
% $\ast$To whom correspondence should be addressed; E-mail: aviv@einstein.yu.edu.
\end{center}

@@

While one can resample a fixed subset of the links, more typically we will be randomly selecting which links to resample with a uniform distribution on the links.  The result of this operation corresponds to averaging our quantity over subsets of links:
\begin{equation}\label{eq:robavgoverlinks}
\langle R (\hbox{stab}, \mu'_{k_1, \ldots, k_m}) \rangle_m =
\frac{1}{\binom{\connectivity}{m}}
\sum_{\{k_1, k_2, \ldots k_m\} \subset \{1, 2,, \ldots \connectivity\}}
R (\hbox{stab}, \mu'_{k_1, \ldots, k_m})
\end{equation}
We are fundamentally interested in the relationship between \ref{eq:condprobgen} and \ref{eq:robavgoverlinks} and topological properties of systems that derive from their associated graphs. In order to investigate this further, we must characterize properties of system graphs in order to evaluate correlations of those properties with dynamical robustness.

@@

\section{Stability and robustness analysis of particular system ensembles}
For systems having two components, we can analytically compute the probability of stability and robustness from \ref{eq:condprobgen}. For those having three components, we can estimate these same quantities using Monte Carlo simulations. Systems of larger size can be analyzed using the symmetry properties of robustness extracted from this analysis. For two-component systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant. Suppose we have a stable matrix
$$
\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}
$$
where $a + c < 0$ and $ac > bd$.  For the case in which $x_1=a,\,x_2=b,\,x_3=c,\,x_4=d$ we need to compute what corresponds to $R(\mathrm{stab},k)$ where $k=1 \ldots 4$. By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent to resampling $d$ so we only need to explicitly compute $R(\mathrm{stab},1)$ and $R(\mathrm{stab},2)$. Suppose that we resample $b$ to compute $R(\mathrm{stab},2)$.
% \begin{strip}
% \begin{align}\label{eq:condprob}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}.
% \end{align}
The denominator of \ref{eq:condprobgen} in this case is given by
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprobgen}
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a' & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

@@

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices, connectivity
$\connectivity_i$, and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $R(\hbox{avg stab} (C_i), 1)$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
R(\hbox{avg stab} (G), 1) =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i R(C_i)}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

@@

Next, we derive an upper bound on the robustness of any graph
constructed from a given set of connected components.  Since
$R(C_i) \le 1$ for all $i$, it follows that increasing $\ell_{ij}$ will
increase $R(G)$.  Given two connected components $C_i$ and $C_j$ with
$v_i$ and $v_j$ nodes respectively, we have a maximum of $n_i n_j$
links going from $C_i$ to $C_j$.  Hence, $\ell_{ij} \le v_i v_j$, so,
we conclude that
\begin{align}
R(G) \le \frac{\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i R(C_i)}
              {\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i},
\end{align}

Since every acyclic digraph can be embedded into a totally ordered
set, we may assume without loss of generality that our components have
been ordered in a way such that, if $(i,j) \in \hier(G)$, then $i <
j$.  Hence, after substituting into our expression and simplifying
using the identity
$$\sum_{i=1}^{n-1}\sum_{j=i+1}^{n}v_i
v_j~=~\frac{1}{2} \left( \sum_{i=1}^{n} v_i \right)^2-\frac{1}{2} \sum_{i=1}^{n}
v_i^2,$$
we arrive at the following upper bound on robustness in terms
of the connected components:
\begin{align} \label{eq:sccmaxrobustness}
R_{\mathrm{max}}(C_1, \ldots C_n) =
\frac{\frac{1}{2} ((\sum_{i=1}^n v_i)^2 - \sum_{i=1}^n v_i^2) +
                    \sum_{i=1}^n \connectivity_i R(\hbox{avg stab}(C_i),1)}
     {\frac{1}{2} ((\sum_{i=1}^n v_i)^2 - \sum_{i=1}^n v_i^2) +
                    \sum_{i=1}^n \connectivity_i}.
\end{align}
Furthermore, this bound is attained.  Suppose that $G_{\mathrm{tot}}$
is the graph on $n$ nodes with a link from node $i$ to node $j$
whenever $i < j$.  Then, by the construction described in the previous
section on network hierachy, we have a graph $G_{\mathrm{max}}$ such
that the components of $G_{\mathrm{max}}$ are $C_1, \ldots C_n$ and
$\hier (G_{\mathrm{max}}) = G_{\mathrm{tot}}$.  By our formula,
$R(\hbox{avg stab} (G_{\mathrm{max}}), 1) = R_{\mathrm{max}}(C_1, \ldots C_n)$.

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Define
\begin{equation*}
M = \left\{(m_0, m_1, \ldots, m_n) \,\bigg|\,
m = \sum_{i=0}^n m_i \quad\&\quad
m_0 \le \sum_{i=1}^{n-1} \sum_{j=i+1}^n \ell_{ij} \quad\&\quad
\forall i \in \{1, \ldots, n\} \, m_i < \connectivity_i \right\}.
\end{equation*}
Then, given $(m_0, m_1, \ldots, m_n) \in M$, there are ${m \choose
m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and
$m_0$ links between strongly connected components.  Hence, our
weighted average becomes
\begin{equation}
R(\hbox{avg stab} (G), m) =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum\limits_{i=1}^n m_i R(\hbox{avg stab} (C_i), m_i) \right)}
     {\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n} m}
\end{equation}

As before, since $R(\hbox{avg stab} (C_i), m_i) \le 1$, we may
increase $R(\hbox{avg stab} (G), m)$ by increasing the maximum
possible value of $m_0$ while keeping the strongly connected
components the same.  Again, if we fix $\hier(G)$, the maximum
possible value of $m_0$ is $\sum_{(i,j) \in \hier(G)} v_i v_j$
whereas, if we allow it to vary, the maximum is $\frac{1}{2} ((\sum_{i=1}^n
v_i)^2 - \sum_{i=1}^n v_i^2)$, which is attained when $\hier(G_{max}) =
G_{tot}$.  Hence, we conclude that $R(\hbox{avg stab} (G), m) \le
R(\hbox{avg stab} (G_{\mathrm{max}}), m)$.

>>>>

% Title must be 150 characters or less
\begin{center}
{\Large
\textbf{Hierarchical Network Structure Promotes Dynamical Robustness}
}
% Insert Author names, affiliations and corresponding author email.
\\[.5cm]
Cameron Smith$^{1}$,
Raymond S. Puzio$^{1}$,
Aviv Bergman$^{1,2,3,4}$,
\\[.5cm]
$^1$Department of Systems and Computational Biology,\\
$^2$Dominick P. Purpura Department of Neuroscience,\\
$^3$Department of Pathology, Albert Einstein College of Medicine,\\
1301 Morris Park Ave, Bronx, NY 10461, USA\\
$^4$Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA
\\[.5cm]
% $\ast$To whom correspondence should be addressed; E-mail: aviv@einstein.yu.edu.
\end{center}

@@

While one can resample a fixed subset of the links, more typically we will be randomly selecting which links to resample with a uniform distribution on the links.  The result of this operation corresponds to averaging our quantity over subsets of links:
\begin{equation}\label{eq:robavgoverlinks}
\langle R (\hbox{stab}, \mu') \rangle_m =
\frac{1}{\binom{\connectivity}{m}}
\sum_{\{k_1, k_2, \ldots k_m\} \subset \{1, 2,, \ldots \connectivity\}}
R (\hbox{stab}, \mu'_{k_1, \ldots, k_m})
\end{equation}
We are fundamentally interested in the relationship between \ref{eq:condprobgen} and \ref{eq:robavgoverlinks} and topological properties of systems that derive from their associated graphs. In order to investigate this further, we must characterize properties of system graphs in order to evaluate correlations of those properties with dynamical robustness.

@@

\section{Stability and robustness analysis of particular system ensembles}
For systems having two components, we can analytically compute the probability of stability and robustness from \ref{eq:condprobgen}. For those having three components, we can estimate these same quantities using Monte Carlo simulations. Systems of larger size can be analyzed using the symmetry properties of robustness extracted from this analysis. For two-component systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant. Suppose we have a stable matrix
$$
\begin{bmatrix}
a & b \\
d & c
\end{bmatrix}
$$
where $a + c < 0$ and $ac > bd$.  For the case in which $x_1=a,\,x_2=b,\,x_3=c,\,x_4=d$ we need to compute what corresponds to $R(\mathrm{stab},\mu'_k)$ where $k=1 \ldots 4$. By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent to resampling $d$ so we only need to explicitly compute $ R(\mathrm{stab},\mu'_1)$ and $R(\mathrm{stab},\mu'_2)$. Suppose that we resample $b$ to compute $R(\mathrm{stab},\mu'_2)$.
% \begin{strip}
% \begin{align}\label{eq:condprob}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}.
% \end{align}
The denominator of \ref{eq:condprobgen} in this case is given by
\begin{align*}
P\left(\textrm{stab } \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprobgen}
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a & b' \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a' & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

@@

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices, connectivity
$\connectivity_i$, and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $\langle R(\hbox{stab} (C_i), \mu') \rangle_1$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
\langle R(\hbox{stab} (G), \mu') \rangle_1 =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i \langle R(\textrm{stab}(C_i),\mu') \rangle_1}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
\langle R(\textrm{stab}(G),\mu') \rangle_1 = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

@@

Next, we derive an upper bound on the robustness of any graph
constructed from a given set of connected components.  Since
$\langle R(\textrm{stab}(C_i), \mu') \rangle_1 \le 1$ for all $i$, it follows that increasing $\ell_{ij}$ will
increase $\langle R(\textrm{stab}(G),\mu') \rangle_1$.  Given two connected components $C_i$ and $C_j$ with
$v_i$ and $v_j$ nodes respectively, we have a maximum of $n_i n_j$
links going from $C_i$ to $C_j$.  Hence, $\ell_{ij} \le v_i v_j$, so,
we conclude that
\begin{align}
\langle R(\textrm{stab}(G),\mu') \rangle_1 \le \frac{\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i \langle R(C_i,\mu') \rangle_1}
              {\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i},
\end{align}

Since every acyclic digraph can be embedded into a totally ordered
set, we may assume without loss of generality that our components have
been ordered in a way such that, if $(i,j) \in \hier(G)$, then $i <
j$.  Hence, after substituting into our expression and simplifying
using the identity
$$\sum_{i=1}^{n-1}\sum_{j=i+1}^{n}v_i
v_j~=~\frac{1}{2} \left( \sum_{i=1}^{n} v_i \right)^2-\frac{1}{2} \sum_{i=1}^{n}
v_i^2,$$
we arrive at the following upper bound on robustness in terms
of the connected components:
\begin{align} \label{eq:sccmaxrobustness}
R_{\mathrm{max}}(C_1, \ldots C_n) =
\frac{\frac{1}{2} ((\sum_{i=1}^n v_i)^2 - \sum_{i=1}^n v_i^2) +
                    \sum_{i=1}^n \connectivity_i \langle R(\hbox{stab}(C_i),\mu') \rangle_1}
     {\frac{1}{2} ((\sum_{i=1}^n v_i)^2 - \sum_{i=1}^n v_i^2) +
                    \sum_{i=1}^n \connectivity_i}.
\end{align}
Furthermore, this bound is attained.  Suppose that $G_{\mathrm{tot}}$
is the graph on $n$ nodes with a link from node $i$ to node $j$
whenever $i < j$.  Then, by the construction described in the previous
section on network hierachy, we have a graph $G_{\mathrm{max}}$ such
that the components of $G_{\mathrm{max}}$ are $C_1, \ldots C_n$ and
$\hier (G_{\mathrm{max}}) = G_{\mathrm{tot}}$.  By our formula,
$\langle R(\hbox{stab} (G_{\mathrm{max}}), \mu') \rangle_1 = R_{\mathrm{max}}(C_1, \ldots C_n)$.

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Define
\begin{equation*}
M = \left\{(m_0, m_1, \ldots, m_n) \,\bigg|\,
m = \sum_{i=0}^n m_i \quad\&\quad
m_0 \le \sum_{i=1}^{n-1} \sum_{j=i+1}^n \ell_{ij} \quad\&\quad
\forall i \in \{1, \ldots, n\} \, m_i < \connectivity_i \right\}.
\end{equation*}
Then, given $(m_0, m_1, \ldots, m_n) \in M$, there are ${m \choose
m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and
$m_0$ links between strongly connected components.  Hence, our
weighted average becomes
\begin{equation}
\langle R(\hbox{stab} (G), \mu') \rangle_m =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum\limits_{i=1}^n m_i \langle R(\hbox{stab} (C_i), \mu') \rangle_{m_i} \right)}
     {\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n} m}
\end{equation}

As before, since $\langle R(\hbox{stab} (C_i), \mu') \rangle_{m_i} \le 1$, we may
increase $\langle R(\hbox{stab} (G), \mu') \rangle_m$ by increasing the maximum
possible value of $m_0$ while keeping the strongly connected
components the same.  Again, if we fix $\hier(G)$, the maximum
possible value of $m_0$ is $\sum_{(i,j) \in \hier(G)} v_i v_j$
whereas, if we allow it to vary, the maximum is $\frac{1}{2} ((\sum_{i=1}^n
v_i)^2 - \sum_{i=1}^n v_i^2)$, which is attained when $\hier(G_{max}) =
G_{tot}$.  Hence, we conclude that $\langle R(\hbox{stab} (G), \mu') \rangle_m \le
\langle R(\hbox{stab} (G_{\mathrm{max}}), \mu') \rangle_m$.
